---
title: "Lab 6 Assignment: Group 21"
author: "Bunia I.A. (s215085), Emma Q.A. (s215090), Joanna S.R. (s215092), Othilia F.S. (s223163) and Tobias A.C. (s215105)"
format:
  html:
    embed-resources: true
editor: visual
---

## Load Libraries

```{r}
packages <- c("tidyverse","ggrepel","broom","patchwork")

for (pck in packages) {
    library(pck, character.only = TRUE)}
```

## Aim

This report aims to analyze biopsy data, using Principal Component Analysis (PCA) to reduce the dimensionality of the dataset. The data is derived from the MASS package, a breast cancer dataset assessing biopsies of breast tumors for 699 patients. Nine attributes were scored on a scale from 1 to 10. Whether the tumor was determined to be benign or malignant is known.

## Load Data

```{r}
#| error: false
#| warning: false
biopsy <- read_csv("https://wilkelab.org/classes/SDS348/data_sets/biopsy.csv")
```

## Data Analysis

#### Exploring the relationship between clump thickness and tumor outcomes.

```{r}
#| error: false
#| warning: false
biopsy |>
  ggplot(data =_) +
  geom_bar(mapping = aes(x = clump_thickness,
                         fill = outcome)) +
  facet_wrap(~outcome) +
  scale_fill_manual(values = c(malignant = "#D55E00", 
                               benign = "#0072B2")) +
  labs(title = "Clump Thickness of Tumors",
       subtitle = "Stratified by outcome",
       x = "Clump Thickness",
       y = "n") +
  theme(legend.position = "")
```

Generally, we observe a correlation between higher clump thickness and an increased likelihood of the tumor being malignant. Furthermore, we see a higher count of benign cases suggesting that, based on the given criteria, benign cases are more prevalent than malignant cases.

#### Exploring the relationship between uniform cell size and uniform cell shape in cancer patients.

```{r}
#| error: false
#| warning: false
biopsy |>
  ggplot(data =_) +
  geom_point(mapping = aes(x = uniform_cell_size,
                          y = uniform_cell_shape,
                          color = outcome)) +
  scale_color_manual(
    values = c(malignant = "#D55E00", 
               benign = "#0072B2")) +
  labs(title = "Cell Shape against Cell size",
       subtitle = "Stratified by outcome",
       color = "Outcome",
       x = "Uniform Cell Size",
       y = "Uniform Cell Shape")
```

From the plot there is a noticeable division in both cell size and cell shape of the benign and malignant tumors. A larger cluster of benign tumor samples, with a few outliers, can be found at lower values of both cell size and shape, indicating that the cells are generally more conserved in shape and size. On the contrary, the range of shape and size for malignant tumor samples are much greater considering that we both observe malignant tumors closer to the values of benign tumors and others almost four times those values. This conjoins with our understanding of cancer cells and that they have greater variation in both size and shape, than "normal" cells.

#### Exploring the relationship between uniform cell size and rate of mitosis in cancer patients.

```{r}
#| error: false
#| warning: false
biopsy |>
  ggplot(data =_) +
  geom_boxplot(mapping = aes(x = uniform_cell_size,
                             y = mitoses,
                             fill = outcome)) +
  scale_fill_manual(values = c(malignant = "#D55E00", 
                               benign = "#0072B2")) +
  labs(title = "Cell Shape against Rate of Mitosis",
       subtitle = "Stratified by outcome",
       fill = "Outcome",
       x = "Uniform Cell Size",
       y = "Mitoses")

```

In our analysis, we noticed a notable disparity in the rate of mitosis between malignant and benign cancer cases. The malignant tumors showcase a greater interquartile range and exhibit prominent outliers, indicating a heightened rate of mitosis. Additionally, the interquartile range illustrates a greater dispersion compared to benign tumors. On the other hand, the boxplot for benign cancer displays more defined upper and lower boundaries, emphasizing a comparatively lower and tightly clustered mitosis rate.

#### Conducting Principal Component Analysis (PCA) to determine the effect of each variable on the outcome.

```{r}
#| error: false
#| warning: false
pca_fit <- biopsy |>
  select(where(is.numeric)) |> # retain only numeric columns
  scale() |> # scale data
  prcomp() # do PCA

pca_fit |>
  augment(biopsy) |> # add original dataset back in
  ggplot(aes(x = .fittedPC1, 
             y = .fittedPC2, 
             color = outcome)) + 
  geom_point(size = 1.5) +
  scale_color_manual(
    values = c(malignant = "#D55E00", 
               benign = "#0072B2")
  ) +
  theme_bw()+
  labs(title = "PC2 vs PC1",
       subtitle = "Stratified via outcome",
       x = "PC1",
       y = "PC2",
       color = "Cancer")
```

From the PCA, we're taking a closer look at the PC1 and PC2 values through a scatter plot. In order to truly interpret the effects of the variables on the cancer being benign or malignant, we also have to visualize the rotation matrix.

#### Plotting a rotation matrix to determine the variance of the variables and the relationship between them.

```{r}
pca_fit |> 
  tidy(matrix = "rotation") |>
  head(10)
```

```{r}
# define arrow style for plotting
arrow_style <- arrow(angle = 20,
                     ends = "first",
                     type = "closed",
                     length = grid::unit(8, "pt")
)

# plot rotation matrix
pca_fit |> 
  tidy(matrix = "rotation") |> 
  pivot_wider(names_from = "PC",
              names_prefix = "PC",
              values_from = "value") |> 
  ggplot(aes(PC1, PC2)) +
  geom_segment(xend = 0,
               yend = 0,
               arrow = arrow_style) +
  geom_label_repel(aes(label = column),
             color = "#D55E00",
             label.padding=0.25,
             size = 2.5,
             min.segment.length = 0.2,
             alpha=0.9) +
  xlim(-1.25, .5) + 
  coord_fixed() + 
  theme_bw() + 
  labs(title = "Rotation Matrix of PC2 vs PC1")
```

In conjunction with the scatter plot of PC2 vs. PC1, we can see that most of the variables move the system towards the left, where most of the cancers are malignant. This means that all of the variables have an effect on whether the cancer is malignant or benign. As it can be observed above, the mitoses variable has the largest influence.

#### Showing the effect of each variable on the outcome.

```{r}
pca_fit |>
  tidy(matrix = "eigenvalues") |>
  ggplot(aes(PC, percent)) +
  geom_col(fill = "#0072B2", alpha = 0.8) +
  scale_x_continuous(breaks = 1:9) +
  scale_y_continuous(
    labels = scales::percent_format(),
    expand = expansion(mult = c(0, 0.01))
  ) +
  theme_bw() +
  labs(title = "Variance Explained by Each PC")
```

It can be observed that the first two PCs generated account for around 80% of the variance. Therefore, by plotting these two together, a graph representing the system's variables' influences well should be achieved.

## Conclusion

The way these biopsies are assessed, show promising results of differentiating between benign and malignant tumors, based on the chosen attributes. After analyzing different attributes, such as clump thickness or uniform cell size, it is possible to see correlations between the assessment values and the outcomes. From the principle component analysis and its various plots, it is seen that PC1 covers above 60% of the variation in the data, which is enough to give an indication as to whether the tumor is benign or malignant.
